How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-07-25 talk/dyla2007/talk.tex
Some reshuffling of the first few slides of the PyPy part.

\begin{frame}
  \frametitle{PyPy's Approach to VM Construction}
  \begin{itemize}
  \item
    Goal: achieve flexibility, simplicity and performance all at one
  \item
    Approach: auto-generate VMs from high-level descriptions of the language
  \item
    ... using meta-programming techniques and aspects (in the original sense)
  \item
    high-level description of the language done by writing an interpreter in
    RPython
  \end{itemize}
  \pause
  \begin{block}{
    What is RPython}
    \begin{itemize}
    \item
      RPython is a subset of Python
    \item
      subset chosen in such a way that type-inference can be performed
    \item
      still a high-level language (unlike SLang or Prescheme)
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Auto-generating VMs}
  \begin{itemize}
  \item
    interpreters written in RPython can be translated to various targets:
    C/Posix, CLR, JVM
  \item
    VMs generated by a flexible translation toolchain (really a compiler)
  \end{itemize}
  \begin{block}{
    Translation Aspects}
    \begin{itemize}
    \item
      the translation toolchain weaves in various aspect into the generated VM
    \item
      examples: GC strategy, threading model
    \item
      early design decisions not necessary
    \item
      non-trivial translation aspect: auto-generating a dynamic compiler from
      the interpreter % XXX maybe we really want to write more about this
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Good Points of the Approach}
  \begin{itemize}
  \item
    dynamic languages can implemented in a high level language
  \item
    doing this separates high-level from low-level concerns
  \item
    generating various VMs from a single source interpreter, which is easy to
    write
  \item
    lots of flexibility at all levels:
    \begin{itemize}
    \item
      when writing the interpreter
    \item
      also possible to adapt the translation toolchain if necessary
    \item
      this allows breaking of abstraction barriers, when necessary (e.g. tagged
      integers)
    \end{itemize}
  \end{itemize}
\end{frame}

>>>>

\begin{frame}
  \frametitle{PyPy's Approach to VM Construction}
  \emph{Goal: achieve flexibility, simplicity and performance together}

  \begin{itemize}
  \item
    Approach: auto-generate VMs from high-level descriptions of the language
  \item
    ... using meta-programming techniques and \emph{aspects}
  \item
    high-level description: an interpreter written in a high-level language
  \item
    ... which we translate (i.e.\ compile) to VMs running on top of various
    targets, like C/Posix, CLR, JVM
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{PyPy}
  \begin{itemize}
  \item
    PyPy = Python interpreter written in RPython + translation toolchain
    for RPython
  \end{itemize}
  \pause
  \begin{block}{
    What is RPython}
    \begin{itemize}
    \item
      RPython is a subset of Python
    \item
      subset chosen in such a way that type-inference can be performed
    \item
      still a high-level language (unlike SLang or Prescheme)
    \item
      ...really a subset, can't give a small example of code that
      doesn't just look like Python :-)
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Auto-generating VMs}
  \begin{itemize}
  \item
    high-level source: early design decisions not necessary
  \item
    we need a custom \emph{translation toolchain} to compile the interpreter
    to a full VM
  \item
    many aspects of the final VM are orthogonal from the interpreter source:
    they are inserted during translation
  \item
    ``monads'', with more ad-hoc control
  \end{itemize}
  \pause
  \begin{block}{
    Examples}
    \begin{itemize}
    \item
      Garbage Collection strategy
    \item
      Threading models (e.g.\ coroutines with CPS...)
    \item
      non-trivial translation aspect: auto-generating a dynamic compiler from
      the interpreter % XXX maybe we really want to write more about this
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Good Points of the Approach}
  \begin{itemize}
  \item
    dynamic languages can be implemented in a high level language
  \item
    doing this separates high-level from low-level concerns
  \item
    generating various VMs from a single source interpreter, which is easy to
    write
  \item
    lots of flexibility at all levels:
    \begin{itemize}
    \item
      when writing the interpreter
    \item
      also possible to adapt the translation toolchain if necessary
    \item
      this allows breaking of abstraction barriers, when necessary (e.g. tagged
      integers)
    \end{itemize}
  \end{itemize}
\end{frame}
